\pkg{cv}: An \proglang{R} Package for Cross-Validation of Regression Models

### 2024-01-03 scratch/notes_john/JSS-article/JSS-article-reduced.tex
added JSS-article-reduced.Rmd (preliminary attempt)

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/mpg-horsepower-scatterplot-1}

@@

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/mpg-horsepower-scatterplot-polynomials-1}

@@

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/mpg-horsepower-MSE-se-1}

@@

Residual standard error: 4.374 on 389 degrees of freedom
Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 44768
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.52622
bias-adjusted cross-validation criterion = 19.49747
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data=Auto)
R> boot::cv.glm(Auto, m.auto.glm)$delta
\end{CodeInput}
\begin{CodeOutput}
[1] 19.24821 19.24787
\end{CodeOutput}
\begin{CodeInput}
R> microbenchmark::microbenchmark(
+   hatvalues = cv(m.auto, k="loo"),
+   Woodbury = cv(m.auto, k="loo", method="Woodbury"),
+   naive = cv(m.auto, k="loo", method="naive"),
+   cv.glm = boot::cv.glm(Auto, m.auto.glm),
+   times=10
+ )
\end{CodeInput}
\begin{CodeOutput}
Warning in microbenchmark::microbenchmark(hatvalues = cv(m.auto, k = "loo"), :
less accurate nanosecond times to avoid potential integer overflows
\end{CodeOutput}
\begin{CodeOutput}
Unit: microseconds
      expr        min        lq       mean     median         uq        max
 hatvalues    977.973    993.43   1143.433   1173.113   1202.981   1296.256
  Woodbury   9932.004  10178.21  11461.993  10363.201  10865.943  17762.266
     naive 215384.152 217809.02 224546.766 219116.505 221670.231 275518.278
    cv.glm 384357.616 388159.59 409612.546 390652.100 441741.913 443226.605
 neval cld
    10 a  
    10 a  
    10  b 
    10   c
\end{CodeOutput}
\end{CodeChunk}

@@

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/polynomial-regression-CV-graph-1}

@@

{\centering \includegraphics[width=0.45\linewidth]{JSS-article_files/figure-latex/polynomial-regression-CV-graph-2-1} \includegraphics[width=0.45\linewidth]{JSS-article_files/figure-latex/polynomial-regression-CV-graph-2-2}

@@

\hypertarget{logistic-regression-for-the-mroz-data}{%
\subsection{Logistic regression for the Mroz
data}\label{logistic-regression-for-the-mroz-data}}

The \code{Mroz} data set from the \pkg{carData} package
\citep[associated with][]{FoxWeisberg:2019} has been used by several
authors to illustrate binary logistic regression; see, in particular
\citet{FoxWeisberg:2019}. The data were originally drawn from the U.S.
Panel Study of Income Dynamics and pertain to married women. Here are a
few cases in the data set:

\begin{CodeChunk}
\begin{CodeInput}
R> data("Mroz", package="carData")
R> head(Mroz, 3)
\end{CodeInput}
\begin{CodeOutput}
  lfp k5 k618 age wc hc       lwg   inc
1 yes  1    0  32 no no 1.2101647 10.91
2 yes  0    2  30 no no 0.3285041 19.50
3 yes  1    3  35 no no 1.5141279 12.04
\end{CodeOutput}
\begin{CodeInput}
R> tail(Mroz, 3)
\end{CodeInput}
\begin{CodeOutput}
    lfp k5 k618 age wc hc       lwg    inc
751  no  0    0  43 no no 0.8881401  9.952
752  no  0    0  60 no no 1.2249736 24.984
753  no  0    3  39 no no 0.8532125 28.363
\end{CodeOutput}
\end{CodeChunk}

The response variable in the logistic regression is \code{lfp},
labor-force participation, a factor coded \code{"yes"} or \code{"no"}.
The remaining variables are predictors:

\begin{itemize}
\tightlist
\item
  \code{k5}, number of children 5 years old of younger in the woman's
  household;
\item
  \code{k618}, number of children between 6 and 18 years old;
\item
  \code{age}, in years;
\item
  \code{wc}, wife's college attendance, \code{"yes"} or \code{"no"};
\item
  \code{hc}, husband's college attendance;
\item
  \code{lwg}, the woman's log wage rate if she is employed, or her
  \emph{imputed} wage rate, if she is not \citep[a variable that][ show
  is problematically defined]{FoxWeisberg:2019}; and
\item
  \code{inc}, family income, in \$1000s, exclusive of wife's income.
\end{itemize}

We use the \code{glm()} function to fit a binary logistic regression to
the \code{Mroz} data:

\begin{CodeChunk}
\begin{CodeInput}
R> m.mroz <- glm(lfp ~ ., data=Mroz, family=binomial)
R> summary(m.mroz)
\end{CodeInput}
\begin{CodeOutput}

Call:
glm(formula = lfp ~ ., family = binomial, data = Mroz)

Coefficients:
             Estimate Std. Error z value Pr(>|z|)
(Intercept)  3.182140   0.644375   4.938 7.88e-07
k5          -1.462913   0.197001  -7.426 1.12e-13
k618        -0.064571   0.068001  -0.950 0.342337
age         -0.062871   0.012783  -4.918 8.73e-07
wcyes        0.807274   0.229980   3.510 0.000448
hcyes        0.111734   0.206040   0.542 0.587618
lwg          0.604693   0.150818   4.009 6.09e-05
inc         -0.034446   0.008208  -4.196 2.71e-05

(Dispersion parameter for binomial family taken to be 1)

    Null deviance: 1029.75  on 752  degrees of freedom
Residual deviance:  905.27  on 745  degrees of freedom
AIC: 921.27

Number of Fisher Scoring iterations: 4
\end{CodeOutput}
\begin{CodeInput}
R> BayesRule(ifelse(Mroz$lfp == "yes", 1, 0), 
+           fitted(m.mroz, type="response"))
\end{CodeInput}
\begin{CodeOutput}
[1] 0.3067729
attr(,"casewise loss")
[1] "y != round(yhat)"
\end{CodeOutput}
\end{CodeChunk}

In addition to the usually summary output for a GLM, we show the result
of applying the \code{BayesRule()} function from the \pkg{cv} package to
predictions derived from the fitted model. Bayes rule, which predicts a
``success'' in a binary regression model when the fitted probability of
success {[}i.e., \(\phi = \Pr(y = 1)\){]} is \(\widehat{\phi} \ge .5\)
and a ``failure'' if \(\widehat{\phi} < .5\).\footnote{\code{BayesRule()}
  does some error checking; \code{BayesRule2()} is similar, but omits
  the error checking, and so can be faster for large problems.} The
first argument to \code{BayesRule()} is the binary \{0, 1\} response,
and the second argument is the predicted probability of success.
\code{BayesRule()} returns the proportion of predictions that are
\emph{in error}, as appropriate for a ``cost'' function.

The value returned by \code{BayesRule()} is associated with an
``attribute'' named \code{"casewise loss"} and set to
\code{"y != round(yhat)"}, signifying that the Bayes rule CV criterion
is computed as the mean of casewise values, here 0 if the prediction for
a case matches the observed value and 1 if it does not (signifying a
prediction error). The \code{mse()} function for numeric responses is
also calculated as a casewise average. Some other criteria, such as the
median absolute error, computed by the \code{medAbsErr()} function in
the \pkg{cv} package, aren't averages of casewise components. The
distinction is important because, to our knowledge, the statistical
theory of cross-validation, for example, in \citet{DavisonHinkley:1997},
\citet{BatesHastieTibshirani:2023}, and \citet{ArlotCelisse:2010}, is
developed for CV criteria like MSE that are means of casewise
components. As a consequence, we limit computation of bias adjustment
and confidence intervals (see below) to criteria that are casewise
averages.

In this example, the fitted logistic regression incorrectly predicts
31\% of the responses; we expect this estimate to be optimistic given
that the model is used to ``predict'' the data to which it is fit.

The \code{"glm"} method for \code{cv()} is largely similar to the
\code{"lm"} method, although the default algorithm, selected explicitly
by \code{method="exact"}, refits the model with each fold removed (and
is thus equivalent to \code{method="naive"} for \code{"lm"} models). For
generalized linear models, \code{method="Woodbury"} or (for LOO CV)
\code{method="hatvalues"} provide approximate results (see the last
section of the vignette for details):

\begin{CodeChunk}
\begin{CodeInput}
R> cv(m.mroz, criterion=BayesRule, seed=248)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 248
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: exact
criterion: BayesRule
cross-validation criterion = 0.3240372
bias-adjusted cross-validation criterion = 0.3195223
95% CI for bias-adjusted CV criterion = (0.2860721, 0.3529724)
full-sample criterion = 0.3067729 
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.mroz, criterion=BayesRule, seed=248, method="Woodbury")
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 248
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: BayesRule
cross-validation criterion = 0.3240372
bias-adjusted cross-validation criterion = 0.3192577
95% CI for bias-adjusted CV criterion = (0.2858076, 0.3527079)
full-sample criterion = 0.3067729 
\end{CodeOutput}
\end{CodeChunk}

To ensure that the two methods use the same 10 folds, we specify the
seed for \proglang{R}'s random-number generator explicitly; here, and as
is common in our experience, the \code{"exact"} and \code{"Woodbury"}
algorithms produce nearly identical results. The CV estimates of
prediction error are slightly higher than the estimate based on all of
the cases.

The printed output includes a 95\% confidence interval for the
bias-adjusted Bayes rule CV criterion.
\citet{BatesHastieTibshirani:2023} show that these confidence intervals
are unreliable for models fit to small samples, and by default
\code{cv()} computes them only when the sample size is 400 or larger and
when the CV criterion employed is an average of casewise components, as
is the case for Bayes rule. See the final section of the vignette for
details of the computation of confidence intervals for bias-adjusted CV
criteria.

Here are results of applying LOO CV to the Mroz model, using both the
exact and the approximate methods:

\begin{CodeChunk}
\begin{CodeInput}
R> cv(m.mroz, k="loo", criterion=BayesRule)
\end{CodeInput}
\begin{CodeOutput}
n-Fold Cross Validation
method: exact
criterion: BayesRule
cross-validation criterion = 0.3200531
bias-adjusted cross-validation criterion = 0.3183001
95% CI for bias-adjusted CV criterion = (0.2849584, 0.3516418)
full-sample criterion = 0.3067729 
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.mroz, k="loo", criterion=BayesRule, method="Woodbury")
\end{CodeInput}
\begin{CodeOutput}
n-Fold Cross Validation
method: Woodbury
criterion: BayesRule
cross-validation criterion = 0.3200531
bias-adjusted cross-validation criterion = 0.3183001
95% CI for bias-adjusted CV criterion = (0.2849584, 0.3516418)
full-sample criterion = 0.3067729 
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.mroz, k="loo", criterion=BayesRule, method="hatvalues")
\end{CodeInput}
\begin{CodeOutput}
n-Fold Cross Validation
method: hatvalues
criterion: BayesRule
cross-validation criterion = 0.3200531
\end{CodeOutput}
\end{CodeChunk}

To the number of decimal digits shown, the three methods produce
identical results for this example.

As for linear models, we report some timings for the various
\textbackslash code\{cv()` methods of computation in LOO CV as well as
for the \code{cv.glm()} function from the \pkg{boot} package (which,
recall, refits the model with each case removed, and thus is comparable
to \code{cv()} with \code{method="exact"}):

\begin{CodeChunk}
\begin{CodeInput}
R> microbenchmark::microbenchmark(
+   hatvalues=cv(m.mroz, k="loo", criterion=BayesRule, method="hatvalues"),
+   Woodbury=cv(m.mroz, k="loo", criterion=BayesRule, method="Woodbury"),
+   exact=cv(m.mroz, k="loo", criterion=BayesRule),
+   cv.glm=boot::cv.glm(Mroz, m.mroz,
+                cost=BayesRule),
+   times=10)
\end{CodeInput}
\begin{CodeOutput}
Unit: milliseconds
      expr         min          lq        mean      median          uq
 hatvalues    1.316141    1.326842    1.654063    1.407571    1.442462
  Woodbury   37.127427   38.641639   39.368192   39.356904   40.562202
     exact 1757.625925 1827.861385 1905.072191 1911.421128 1951.049944
    cv.glm 2062.693354 2122.649130 2149.761134 2140.079440 2184.788525
         max neval cld
    4.022428    10 a  
   41.060188    10 a  
 2128.307540    10  b 
 2299.544573    10   c
\end{CodeOutput}
\end{CodeChunk}

There is a substantial time penalty associated with exact computations.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> library(nlme)
R> hsb.lme <- lme(mathach ~ mean.ses*cses + sector*cses,
+                  random = ~ cses | school, data=HSB,
+                control=list(opt="optim"))
R> summary(hsb.lme)
\end{CodeInput}
\begin{CodeOutput}
Linear mixed-effects model fit by REML
  Data: HSB 
       AIC      BIC    logLik
  46524.78 46593.57 -23252.39

@@

We start by generating the data, using parameters consistent with the
description above and meant to highlight the issues that arise in
cross-validating mixed-effects models:\footnote{The \code{regsubsets()}
  function computes several measures of model predictive performance,
  including the \(R^2\) and \(R^2\) adjusted for degrees of freedom, the
  residual sums of squares, Mallows's \(C_p\), and the BIC. Several of
  these are suitable for comparing models with differing numbers of
  coefficients---we use the BIC below---but all necessarily agree when
  comparing models with the \emph{same} number of coefficients.}

@@

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/plot1-1}

@@

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/plot-fits-mod0-1}

@@

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/plot-fits-mod1-1}

@@

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/plot-fits-mod2-1}

@@

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/plot-fits-mod3-1}

@@

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/cross-validation-clusters-1}

@@

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/cross-validation-cases-1}

@@

\hypertarget{example-crossed-random-effects}{%
\subsection{Example: Crossed random
effects}\label{example-crossed-random-effects}}

Crossed random effects arise when the structure of the data aren't
strictly hierarchical. Nevertheless, crossed and nested random effects
can be handled in much the same manner, by refitting the mixed-effects
model to the data with a fold of clusters or cases removed and using the
refitted model to predict the response in the removed fold.

We'll illustrate with data on pig growth, introduced by \citet[Table
3.1]{DiggleLiangZeger:1994}. The data are in the \code{Pigs} data frame
in the \pkg{cv} package:

\begin{CodeChunk}
\begin{CodeInput}
R> head(Pigs, 9)
\end{CodeInput}
\begin{CodeOutput}
  id week weight
1  1    1   24.0
2  1    2   32.0
3  1    3   39.0
4  1    4   42.5
5  1    5   48.0
6  1    6   54.5
7  1    7   61.0
8  1    8   65.0
9  1    9   72.0
\end{CodeOutput}
\begin{CodeInput}
R> head(xtabs(~ id + week, data=Pigs), 3)
\end{CodeInput}
\begin{CodeOutput}
   week
id  1 2 3 4 5 6 7 8 9
  1 1 1 1 1 1 1 1 1 1
  2 1 1 1 1 1 1 1 1 1
  3 1 1 1 1 1 1 1 1 1
\end{CodeOutput}
\begin{CodeInput}
R> tail(xtabs(~ id + week, data=Pigs), 3)
\end{CodeInput}
\begin{CodeOutput}
    week
id   1 2 3 4 5 6 7 8 9
  46 1 1 1 1 1 1 1 1 1
  47 1 1 1 1 1 1 1 1 1
  48 1 1 1 1 1 1 1 1 1
\end{CodeOutput}
\end{CodeChunk}

Each of 48 pigs is observed weekly over a period of 9 weeks, with the
weight of the pig recorded in kg. The data are in ``long'' format, as is
appropriate for use with the \code{lmer()} function in the \pkg{lme4}
package. The data are very regular, with no missing cases.

The following graph, showing the growth trajectories of the pigs, is
similar to Figure 3.1 in \citet{DiggleLiangZeger:1994}; we add an
overall least-squares line and a loess smooth, which are nearly
indistinguishable:

\begin{CodeChunk}
\begin{CodeInput}
R> plot(weight ~ week, data=Pigs, type="n")
R> for (i in unique(Pigs$id)){
+   with(Pigs, lines(x=1:9, y=Pigs[id == i, "weight"],
+                    col="gray"))
+ }
R> abline(lm(weight ~ week, data=Pigs), col="blue", lwd=2)
R> lines(with(Pigs, loess.smooth(week, weight, span=0.5)),
+       col="magenta", lty=2, lwd=2)
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=0.6\linewidth]{JSS-article_files/figure-latex/pigs-graph-1}

\caption[Growth trajectories for 48 pigs, with overall least-squares line (sold blue) and loess line (broken magenta)]{Growth trajectories for 48 pigs, with overall least-squares line (sold blue) and loess line (broken magenta).}\label{fig:pigs-graph}
\end{figure}
\end{CodeChunk}

The individual ``growth curves'' and the overall trend are generally
linear, with some tendency for variability of pig weight to increase
over weeks (a feature of the data that we ignore in the mixed model that
we fit to the data below).

The \proglang{Stata} mixed-effects models manual proposes a model with
crossed random effects for the \code{Pigs} data \citep[ page
37]{Stata:2023}:

\begin{quote}
{[}S{]}uppose that we wish to fit \[
\mathrm{weight}_{ij} = \beta_0 + \beta_1 \mathrm{week}_{ij} + u_i + v_j + \varepsilon_{ij} 
\] for the \(i = 1, \ldots, 9\) weeks and \(j = 1, \dots, 48\) pigs and
\[
u_i \sim N(0, \sigma^2_u); v_j \sim N(0, \sigma^2_v ); \varepsilon_{ij} \sim N(0, \sigma^2_\varepsilon)
\] all independently. That is, we assume an overall population-average
growth curve \(\beta_0 + \beta_1 \mathrm{week}\) and a random
pig-specific shift. In other words, the effect due to week, \(u_i\), is
systematic to that week and common to all pigs. The rationale behind
{[}this model{]} could be that, assuming that the pigs were measured
contemporaneously, we might be concerned that week-specific random
factors such as weather and feeding patterns had significant systematic
effects on all pigs.
\end{quote}

Although we might prefer an alternative model,\footnote{These are
  repeated-measures data, which would be more conventionally modeled
  with autocorrelated errors within pigs. The \code{lme()} function in
  the \pkg{nlme} package, for example, is capable of fitting a
  mixed-model of this form.} we think that this is a reasonable
specification.

The \proglang{Stata} manual fits the mixed model by maximum likelihood
(rather than REML), and we duplicate the results reported there using
\code{lmer()}:

\begin{CodeChunk}
\begin{CodeInput}
R> m.p <- lmer(weight ~ week + (1 | id) + (1 | week),
+             data=Pigs, REML=FALSE, # i.e., ML
+             control=lmerControl(optimizer="bobyqa"))
R> summary(m.p)
\end{CodeInput}
\begin{CodeOutput}
Linear mixed model fit by maximum likelihood  ['lmerMod']
Formula: weight ~ week + (1 | id) + (1 | week)
   Data: Pigs
Control: lmerControl(optimizer = "bobyqa")

     AIC      BIC   logLik deviance df.resid 
  2037.6   2058.0  -1013.8   2027.6      427

Scaled residuals: 
    Min      1Q  Median      3Q     Max 
-3.7750 -0.5418  0.0054  0.4762  3.9816

Random effects:
 Groups   Name        Variance Std.Dev.
 id       (Intercept) 14.83622 3.8518  
 week     (Intercept)  0.08499 0.2915  
 Residual              4.29733 2.0730  
Number of obs: 432, groups:  id, 48; week, 9

Fixed effects:
            Estimate Std. Error t value
(Intercept) 19.35561    0.63340   30.56
week         6.20990    0.05393  115.14

Correlation of Fixed Effects:
     (Intr)
week -0.426
\end{CodeOutput}
\end{CodeChunk}

We opt for the non-default \code{"bobyqa"} optimizer because it provides
more numerically stable results for subsequent cross-validation in this
example.

We can then cross-validate the model by omitting folds composed of pigs,
folds composed of weeks, or folds composed of pig-weeks (which in the
\code{Pigs} data set correspond to individual cases, using only the
fixed effects):

\begin{CodeChunk}
\begin{CodeInput}
R> cv(m.p, clusterVariables="id")
\end{CodeInput}
\begin{CodeOutput}
n-Fold Cross Validation based on 48 {id} clusters
cross-validation criterion = 19.97309
bias-adjusted cross-validation criterion = 19.96496
95% CI for bias-adjusted CV criterion = (17.1249, 22.80502)
full-sample criterion = 19.20076 
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.p, clusterVariables="week")
\end{CodeInput}
\begin{CodeOutput}
boundary (singular) fit: see help('isSingular')
\end{CodeOutput}
\begin{CodeOutput}
n-Fold Cross Validation based on 9 {week} clusters
cross-validation criterion = 19.31219
bias-adjusted cross-validation criterion = 19.30499
95% CI for bias-adjusted CV criterion = (16.56624, 22.04373)
full-sample criterion = 19.20076 
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.p, clusterVariables=c("id", "week"), k=10, seed=8469)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 8469
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation based on 432 {id, week} clusters
cross-validation criterion = 19.2352
bias-adjusted cross-validation criterion = 19.23294
95% CI for bias-adjusted CV criterion = (16.49263, 21.97325)
full-sample criterion = 19.20076 
\end{CodeOutput}
\end{CodeChunk}

We can also cross-validate the individual cases taking account of the
random effects (employing the same 10 folds):

\begin{CodeChunk}
\begin{CodeInput}
R> cv(m.p, k=10, seed=8469)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 8469
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
cross-validation criterion = 5.15835
bias-adjusted cross-validation criterion = 5.072933
95% CI for bias-adjusted CV criterion = (4.122952, 6.022913)
full-sample criterion = 3.795972 
\end{CodeOutput}
\end{CodeChunk}

Because these predictions are based on BLUPs, they are more accurate
than the predictions based only on fixed effects.\footnote{Even though
  there is only one observation per combination of pigs and weeks, we
  can use the BLUP for the omitted case because of the crossed structure
  of the random effects; that is each pig-week has a pig random effect
  and a week random effect. Although it probably isn't sensible, we can
  imagine a mixed model for the pig data that employs nested random
  effects, which would be specified by
  \code{lmer(weight ~ week + (1 | id/week), data=Pigs)}---that is, a
  random intercept that varies by combinations of \code{id} (pig) and
  \code{week}. This model can't be fit, however: With only one case per
  combination of \code{id} and \code{week}, the nested random-effect
  variance is indistinguishable from the case-level variance.} As well,
the difference between the MSE computed for the model fit to the full
data and the CV estimates of the MSE is greater here than for
cluster-based predictions.

\hypertarget{replicating-cross-validation}{%
\section{Replicating
cross-validation}\label{replicating-cross-validation}}

Assuming that the number of cases \(n\) is a multiple of the number of
folds \(k\)---a slightly simplifying assumption---the number of possible
partitions of cases into folds is \(\frac{n!}{[(n/k)!]^k}\), a number
that grows very large very quickly. For example, for \(n = 10\) and
\(k = 5\), so that the folds are each of size \(n/k = 2\), there are
\(113,400\) possible partitions; for \(n=100\) and \(k=5\), where
\(n/k = 20\), still a small problem, the number of possible partitions
is truly astronomical, \(1.09\times 10^{66}\).

Because the partition into folds that's employed is selected randomly,
the resulting CV criterion estimates are subject to sampling error. (An
exception is LOO cross-validation, which is not at all random.) To get a
sense of the magnitude of the sampling error, we can repeat the CV
procedure with different randomly selected partitions into folds. All of
the CV functions in the \pkg{cv} package are capable of repeated
cross-validation, with the number of repetitions controlled by the
\code{reps} argument, which defaults to \code{1}.

Here, for example, is 10-fold CV for the Mroz logistic regression,
repeated 5 times:

\begin{CodeChunk}
\begin{CodeInput}
R> cv(m.mroz, criterion=BayesRule, seed=248, reps=5, 
+    method="Woodbury")
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 248
\end{CodeOutput}
\begin{CodeOutput}
R RNG seed set to 68134
\end{CodeOutput}
\begin{CodeOutput}
R RNG seed set to 767359
\end{CodeOutput}
\begin{CodeOutput}
R RNG seed set to 556270
\end{CodeOutput}
\begin{CodeOutput}
R RNG seed set to 882966
\end{CodeOutput}
\begin{CodeOutput}

Replicate 1:
10-Fold Cross Validation
method: Woodbury
criterion: BayesRule
cross-validation criterion = 0.3200531
bias-adjusted cross-validation criterion = 0.3130109
95% CI for bias-adjusted CV criterion = (0.2796692, 0.3463526)
full-sample criterion = 0.3067729

Replicate 2:
10-Fold Cross Validation
method: Woodbury
criterion: BayesRule
cross-validation criterion = 0.3160691
bias-adjusted cross-validation criterion = 0.311697
95% CI for bias-adjusted CV criterion = (0.2784665, 0.3449274)
full-sample criterion = 0.3067729

Replicate 3:
10-Fold Cross Validation
method: Woodbury
criterion: BayesRule
cross-validation criterion = 0.314741
bias-adjusted cross-validation criterion = 0.3086212
95% CI for bias-adjusted CV criterion = (0.2754285, 0.341814)
full-sample criterion = 0.3067729

Replicate 4:
10-Fold Cross Validation
method: Woodbury
criterion: BayesRule
cross-validation criterion = 0.3240372
bias-adjusted cross-validation criterion = 0.318069
95% CI for bias-adjusted CV criterion = (0.2846189, 0.3515192)
full-sample criterion = 0.3067729

Replicate 5:
10-Fold Cross Validation
method: Woodbury
criterion: BayesRule
cross-validation criterion = 0.3240372
bias-adjusted cross-validation criterion = 0.3192577
95% CI for bias-adjusted CV criterion = (0.2858076, 0.3527079)
full-sample criterion = 0.3067729

Average:
10-Fold Cross Validation
method: Woodbury
criterion: BayesRule
cross-validation criterion = 0.3198318 (0.003886968)
bias-adjusted cross-validation criterion = 0.3139445 (0.004009339)
full-sample criterion = 0.3067729 
\end{CodeOutput}
\end{CodeChunk}

When \code{reps} \textgreater{} \code{1}, the result returned by
\code{cv()} is an object of class \code{"cvList"}---literally a list of
\code{"cv"} objects. The results are reported for each repetition and
then averaged across repetitions, with the standard deviations of the CV
criterion and the biased-adjusted CV criterion given in parentheses. In
this example, there is therefore little variation across repetitions,
increasing our confidence in the reliability of the results.

Notice that the seed that's set in the \code{cv()} command pertains to
the first repetition and the seeds for the remaining repetitions are
then selected pseudo-randomly.\footnote{Because of the manner in which
  the computation is performed, the order of the replicates in the
  \code{"cvList"} object returned by \code{cv()} isn't the same as the
  order in which the replicates are computed. Each element of the
  result, however, is a \code{"cv"} object with the correct
  random-number seed saved, and so this technical detail can be safely
  ignored. The individual \code{"cv"} objects are printed in the order
  in which they are stored rather than the order in which they are
  computed.} Setting the first seed, however, makes the entire process
easily replicable, and the seed for each repetition is stored in the
corresponding element of the \code{"cvList"} object (which isn't,
however, saved in the example).

It's also possible to replicate CV when comparing competing models via
the \code{cv()} method for \code{"modList"} objects. Recall our
comparison of polynomial regressions of varying degree fit to the
\code{Auto} data; we performed 10-fold CV for each of 10 models. Here,
we replicate that process 5 times for each model and graph the results:

\begin{CodeChunk}
\begin{CodeInput}
R> cv.auto.reps <- cv(models(m.1, m.2, m.3, m.4, m.5,
+                         m.6, m.7, m.8, m.9, m.10),
+                  data=Auto, seed=8004, reps=5)
R> plot(cv.auto.reps)
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/model-comparison-with-reps-1}

\caption[ Replicated cross-validated 10-fold CV as a function of polynomial degree, $p$]{ Replicated cross-validated 10-fold CV as a function of polynomial degree, $p$}\label{fig:model-comparison-with-reps}
\end{figure}
\end{CodeChunk}

The graph shows both the average CV criterion and its range for each of
the competing models.

\hypertarget{cross-validating-model-selection}{%
\section{Cross-validating model
selection}\label{cross-validating-model-selection}}

@@

\hypertarget{mrozs-logistic-regression-revisited}{%
\subsection{Mroz's logistic regression
revisited}\label{mrozs-logistic-regression-revisited}}

For a contrasting example we apply model selection to Mroz's logistic
regression for married women's labor-force participation. First, recall
the logistic regression model that we fit to the \code{Mroz} data:

\begin{CodeChunk}
\begin{CodeInput}
R> summary(m.mroz)
\end{CodeInput}
\begin{CodeOutput}

Call:
glm(formula = lfp ~ ., family = binomial, data = Mroz)

>>>>

{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/mpg-horsepower-scatterplot-1}

@@

{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/mpg-horsepower-scatterplot-polynomials-1}

@@

{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/mpg-horsepower-MSE-se-1}

@@

Residual standard error: 4.374 on 389 degrees of freedom
Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 423836
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.26863
bias-adjusted cross-validation criterion = 19.25351
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data=Auto)
R> boot::cv.glm(Auto, m.auto.glm)$delta
\end{CodeInput}
\begin{CodeOutput}
[1] 19.24821 19.24787
\end{CodeOutput}
\begin{CodeInput}
R> microbenchmark::microbenchmark(
+   hatvalues = cv(m.auto, k="loo"),
+   Woodbury = cv(m.auto, k="loo", method="Woodbury"),
+   naive = cv(m.auto, k="loo", method="naive"),
+   cv.glm = boot::cv.glm(Auto, m.auto.glm),
+   times=10
+ )
\end{CodeInput}
\begin{CodeOutput}
Warning in microbenchmark::microbenchmark(hatvalues = cv(m.auto, k = "loo"), :
less accurate nanosecond times to avoid potential integer overflows
\end{CodeOutput}
\begin{CodeOutput}
Unit: microseconds
      expr        min         lq       mean     median         uq        max
 hatvalues    984.287   1153.412   1160.394   1189.902   1199.414   1285.104
  Woodbury  10145.122  10213.592  10525.532  10463.385  10657.581  11476.351
     naive 216360.403 217763.218 223882.308 218184.226 219846.551 273572.951
    cv.glm 380361.674 382182.689 400581.980 386284.739 436401.540 439866.368
 neval cld
    10 a  
    10 a  
    10  b 
    10   c
\end{CodeOutput}
\end{CodeChunk}

@@

{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/polynomial-regression-CV-graph-1}

@@

{\centering \includegraphics[width=0.45\linewidth]{JSS-article-reduced_files/figure-latex/polynomial-regression-CV-graph-2-1} \includegraphics[width=0.45\linewidth]{JSS-article-reduced_files/figure-latex/polynomial-regression-CV-graph-2-2}

@@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> library(nlme)
\end{CodeInput}
\begin{CodeOutput}

Attaching package: 'nlme'
\end{CodeOutput}
\begin{CodeOutput}
The following object is masked from 'package:lme4':

    lmList
\end{CodeOutput}
\begin{CodeOutput}
The following object is masked from 'package:dplyr':

    collapse
\end{CodeOutput}
\begin{CodeInput}
R> hsb.lme <- lme(mathach ~ mean.ses*cses + sector*cses,
+                  random = ~ cses | school, data=HSB,
+                control=list(opt="optim"))
R> summary(hsb.lme)
\end{CodeInput}
\begin{CodeOutput}
Linear mixed-effects model fit by REML
  Data: HSB 
       AIC      BIC    logLik
  46524.78 46593.57 -23252.39

@@

We start by generating the data, using parameters consistent with the
description above and meant to highlight the issues that arise in
cross-validating mixed-effects models:\footnote{We invite the interested
  reader to experiment with varying the parameters of our example.}

@@

{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/plot1-1}

@@

{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/plot-fits-mod0-1}

@@

{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/plot-fits-mod1-1}

@@

{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/plot-fits-mod2-1}

@@

{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/plot-fits-mod3-1}

@@

{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/cross-validation-clusters-1}

@@

{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/cross-validation-cases-1}

@@

@@
